\documentclass{article}
\usepackage{enumitem}
\newtheorem{example}{Example}
\begin{document}

\title{Fall 2023 B561 Assignment 1 \\
Expressing Queries and Constraints in SQL and in Tuple Relational Calculus (TRC)}
\author{Pavan Kalyan Thota, Nihanth Vuddanti}
\date{Released: September 1, 2023\\ Due: September 14, 2023 }
\maketitle


\section{Introduction}

The goals for this assignment are to 
\begin{enumerate}
\item become familiar with the PostgreSQL system\footnote{To solve this assignment, you will need to download and install PostgreSQL (version 12 or higher) on your computer.};
\item create a relational database and populate it with data;
\item examine the side-effects on the state of the database caused by inserts and deletes in the presence or absence
of primary and foreign key constraints;
\item formulate some queries in SQL and evaluate them in PostgreSQL; and
\item translate TRC queries to SQL and 
formulate queries and constraints in TRC.\footnote{To solve problems related to TRC, follow the syntax and semantics described in the {\tt TRC\_SQL.pdf} document
in the module \emph{Tuple Relational Calculus and SQL (lecture 4)}.   That document contains multiple examples of
TRC queries and constraints and how they can be translated to SQL.}
\end{enumerate}

To turn in your assignment, you will need to upload to Canvas a single file with name {\tt assignment1.sql} which contains 
the necessary SQL statements that solve the problems in this assignment.   
The {\tt assignment1.sql} file must be such that the AI's can run it in their PostgreSQL environment.  
In addition, you will need to upload a separate {\tt assignment1.txt} file that contains the results of running
your queries.
We have posted the exact requirements and an example for uploading your solution files.  (See the module
{\tt Instructions for turning in assignments}.)
Finally, you will need to upload an {\tt assignment1.pdf} file that contains the solutions for problems related to TRC.\footnote{It is strongly recommended that you use Latex to write TRC formulas and queries.
For a good way to learn about Latex, look at 
%\begin{verbatim}
https://www.overleaf.com/learn/latex/Free\_online\_introduction\_to\_LaTeX\_(part\_1).
%\end{verbatim}
You can also inspect the Latex source code for this assignment as well as the document TRC\_SQL.tex provided in Module 4.}



For the problems in this assignment we will use the following database schema:\footnote{The primary key, which may consist of one or more attributes, of each of these relations is underlined.}

\begin{center}
{\tt
  \begin{tabular}{l}
  {Person}($\underline{\tt pid}$, pname, city) \\
  {Company}($\underline{\tt cname}, {\tt headquarter}$) \\
  {Skill}($\underline{\tt skill}$) \\
  {worksFor}($\underline{\tt pid}$, cname, salary) \\
  {companyLocation}($\underline{\tt cname, city}$) \\
  {personSkill}($\underline{\tt pid, skill}$) \\
  {hasManager}($\underline{\tt eid, mid}$) \\
  {Knows}($\underline{\tt pid1, pid2}$) \\
  \end{tabular}
  }
\end{center}

In this database we maintain a set of persons ({\tt Person}), a set
of companies ({\tt Company}), and a set of (job) skills ({\tt Skill}).  
The {\tt pname} attribute in {\tt Person} is the name of the person.  
The {\tt city} attribute in {\tt Person} specifies the city in which the person lives.  
The {\tt cname} attribute in {\tt Company} is the name of the company.
The {\tt headquarter} attribute in {\tt Company} is the name of the city wherein the company has its headquarter.
The {\tt skill} attribute in {\tt Skill} is the name of a (job) skill.

A person can work for at most one company. This information is maintained in the {\tt worksFor} relation. (We permit that a person does not work for any company.)
The {\tt salary} attribute in {\tt worksFor} specifies the salary made by the person.

The {\tt city} attribute in {\tt companyLocation} indicates a city in which the company is located.
(Companies may be located in multiple cities.)

A person can have multiple job skills. This information is maintained in the {\tt personSkill} relation.  A job skill can be
the job skill of multiple persons.  (A person may not have any job skills, and a job skill may
have no persons with that skill.)

A pair $(e,m)$ in {\tt hasManager} indicates that person $e$ has  
person $m$ as one of his or her managers.
We permit that an employee has multiple managers and that a manager  may manage
multiple employees.  (It is possible that an employee has no manager
and that an employee is not a manager.)
We further require that 
an employee and his or her managers must work for the
same company.

The relation {\tt Knows} maintains a set of pairs $(p_1,p_2)$ 
where $p_1$ 
and $p_2$ are pids of persons.   The pair $(p_1,p_2)$ indicates 
that the person with
pid $p_1$ knows the person with pid $p_2$.
We do not assume that the relation {\tt Knows} is
symmetric: it is possible that $(p_1,p_2)$ is in the relation but that $(p_2,p_1)$ is not.

The domain for the attributes {\tt pid}, {\tt salary}, {\tt eid}, and {\tt mid} is {\tt integer}.   The domain for all other attributes is {\tt text}.

We assume the following foreign key constraints:
\begin{itemize}
\item {\tt pid} is a foreign key in {\tt worksFor} referencing the primary key {\tt pid} in {\tt Person};
\item {\tt cname} is a foreign key in {\tt worksFor} referencing the primary key {\tt cname} in {\tt Company};
\item {\tt cname} is a foreign key in {\tt companyLocation} referencing the primary key {\tt cname} in {\tt Company};
\item {\tt pid} is a foreign key in {\tt personSkill} referencing the primary key {\tt pid} in {\tt Person};
\item {\tt skill} is a foreign key in {\tt personSkill} referencing the primary key {\tt skill} in {\tt Skill};
\item {\tt eid} is a foreign key in {\tt hasManager} referencing the primary key {\tt pid} in {\tt Person}; and
\item {\tt mid} is a foreign key in {\tt hasManager} referencing the primary key {\tt pid} in {\tt Person};
\item {\tt pid1} is a foreign key in {\tt Knows} referencing the 
primary key {\tt pid} in {\tt Person}; and
\item {\tt pid2} is a foreign key in {\tt Knows} referencing the 
primary key {\tt pid} in {\tt Person}

\end{itemize}

The file {\tt data.sql} contains the data supplied for this assignment.

\newpage
\section{Database creation and impact of constraints on insert and delete statements.}

Create a database in PostgreSQL that stores the data provided in the \\{\tt Assignment1Script.sql} file. Make sure to specify primary and foreign keys. 

\begin{enumerate}
\item Provide 4 conceptually different examples that illustrate how the
presence or absence of primary and foreign keys affect insert and
deletes in these relations.  To solve this problem, you will need to
experiment with the relation schemas and instances for this
assignment.  For example, you should consider altering primary keys
and foreign key constraints and then consider various sequences of
insert and delete operations.  You may need to change some of the
relation instances to observe the desired effects.  Certain inserts
and deletes should succeed but other should generate error
conditions.  (Consider the lecture notes about keys, foreign keys,
and inserts and deletes as a guide to solve this problem.) (5 Points)
  \end{enumerate}
 \newpage
  \section{Formulating queries in SQL}\label{SQLQueries}

For this assignment,  you are required to 
use  tuple variables in your SQL statements.   
For example, in formulating the query ``Find the pid and pname of each person who lives in Bloomington" you should write the query

\begin{center}{\tt
\begin{tabular}{ll}
SELECT &p.pid, p.pname\\
FROM & Person p\\
WHERE & p.city = `Bloomington'
\end{tabular}}
\end{center}
rather than
\begin{center}{\tt
\begin{tabular}{ll}
SELECT &pid, pname\\
FROM & Person\\
WHERE & city = `Bloomington' 
\end{tabular}}
\end{center}

Write SQL statements for the following queries.   
Make sure that each of your queries returns a set but not a bag.  In other words, make appropriate use of the
{\tt DISTINCT} clause where necessary.

 You can {\bf not} use the SQL JOIN operations or SQL aggregate functions
such as {\tt COUNT}, {\tt SUM}, {\tt MAX}, {\tt MIN}, etc in your solutions.
\begin{enumerate}[resume]
\item Find the pid, pname of each person who (a) lives in MountainView, (b) works for a company which is headquartered in MountainView , and (c) has a salary less than or equal to 60000. (5 Points)

\item Find the name of each company with atleast 2 employees. (5 Points)

\item Find the ID of each person who knows someone living in Bloomington and having Networks skill ( If there is a person p1, we print p1 if he/she knows atleast one person p2 such that p2 lives in Bloomington and has the Networks skill ). (5 Points)

\item Find the pairs of pids $(p_1, p_2)$ such that p1 and p2 both have atleast one common skill. (5 Points)

\item Find the ID, name, and salary of each person who works for company located in Cupertino and who earns the second lowest salary  at that company.(5 Points)

\item Find the ID of managers with AI skill who are managing atleast 2 people. (5 Points)

\item Find the name of each company that has atleast two employees along with lowest salary made by an employee in that company. (The query is returns the company name and the lowest salary made by any of its employee as output). (5 Points)

\item 
Find the Id, Name and City pf a person who knows atleast one person who knows another person that earns more than 65000.\\
Let us consider 3 people p1,p2 and p3. p1 knows p2 and p2 knows p3. p3 earns more than 65000. The query returns the ID,name and city of p1. (5 Points)

\newpage
% \item Find the pairs $(c_1, c_2)$ of names of companies whose headquarters are located in the same city.
% \item Find the pid and pname of each person who lives in a city that is different than each city in which his or her managers live.
% (Persons who have no manager should not be included in the answer.)
% \item Find each skill that is the skill of at most 2 persons.
% \item Find the pid, pname, and salary of each employee who has at least two managers such that these managers have a common job skill but provided that it is not the `Networks' skill.
% \item Find the cname of each company that not only employs persons                                                                      
% who live in MountainView. 
% (In other words, there exists at least one employee of such a company
% who does not live in MountainView.)
% \item For each company, list its name along with the highest salary made by employees who work for it.
% \item Find the pid and pname of each employee who has a salary that is higher than the salary of each of his or her managers.
% (Employees who have no manager should not be included.)

\end{enumerate}

\section{Translating TRC queries to SQL}

Consider the following queries formulated in TRC.   Translate each of these queries to an equivalent SQL query.\footnote{
You can not use SQL {\tt JOIN} operations or
aggregate functions.}

This underscores the close correspondence between TRC and SQL.

The SQL queries should be included in the {\tt assignment1.sql} file and their outputs should be reported in the
{\tt assignment.txt} file.

\begin{enumerate}[resume]
\item 
{\small
\[
\begin{array}{ll}
\{p.pid, p.pname, w.cname, w.salary\mid Person(p) \land worksFor(w)\land Company(c)\land p.pid = w.pid\\
\qquad p.city = \mbox{\rm `Bloomington'}\, \land\, 40000 \leq w.salary\, \land\, w.cname = c.cname\land c.headquarter = \mbox{\rm 'Seattle'}\}.\\ (5 Points)
\end{array}
\]
}


\item 
{\small
\[
\begin{array}{ll}
\{ps.pid \mid personSKill(ps) \land hasManager(hm) \land Person(m) \land ps.skill = \mbox{\rm 'Programming'} \\ \quad  \land hm.eid = ps.pid \land hm.mid = m.pid \land \exists ms (psersonSkill(ms) \land ms.pid = m.pid \\ \quad \land ms.skill != \mbox{\rm 'Programming'})\}. (5 Points)
\end{array}
\]
}

\item 
{\small
\[
\begin{array}{ll}
\{ps.skill\mid Person(p) \land personSkill(ps)\land  p.pid = ps.pid \land p.city != \mbox{`Bloomington'} \\
\land \exists m (Person(m) \land hasManager(hm) \land hm.eid = p.pid \land hm.mid = m.pid \land m.city = \mbox{`Bloomington'} ) \}.\\ (5 Points)
\end{array}
\]
}


\item 
{\small
\[
\begin{array}{ll}
\{m.pid, m.pname\mid Person(m) \land \\
\qquad \forall hm ((hasManager(hm)\land hm.mid = m.pid)\rightarrow 
\exists e (Person(e)\land hm.eid = e.pid \land e.city = m.city))\}\\(5 Points)
\end{array}
\]
}

\end{enumerate}
\newpage
\section{Formulating queries in the Tuple Relational Calculus}


The solutions of these problems should be included in the {\tt assignment1.pdf} file.

\begin{enumerate}[resume]
\item Find the pid, pname of each person who (a) lives in Bloomington, (b) works for a company where he or she earn a salary that is
in range [30000, 50000] (inclusive) (c) has at least one manager. It should be a single query. (5 Points)

\item Find the pid and pname of each person who lives in a city
that is different than each city in which his or her managers live. (Persons
who have no manager should also be included in the answer.) (5 Points)

\item Find the pid, pname, and salary of each employee who has
at least two managers such that these managers have a common job skill
but provided that it is not the `Programming' skill. (5 Points)

\item For each company, list its name along with the highest salary
made by employees who work for it. (5 Points)

\end{enumerate}
\newpage

\section{Formulating constraints in the Tuple Relational Calculus}

Formulate the following constraints in TRC and as boolean SQL queries.    

The TRC solutions of these problems should be included in the {\tt assignment1.pdf} file and
the SQL solutions should be included in the {\tt assignment1.sql} file.

Here is an example of what is expected for your answers.

\begin{example}
Consider the constraint ``\emph{Each skill is the skill of a person.}''
In TRC, this constraint can be formulated as follows:
$$
\begin{array}{l}
\forall s\ Skill(s) \rightarrow \exists ps\, (personSkill(ps) \land ps.skill = s.skill)
\end{array}
$$
or, alternatively
$$
\begin{array}{l}
\lnot \exists s (Skill(s) \land \lnot \exists ps (personSkill(ps) \land ps.skill = s.skill)).
\end{array}
$$
This constraint can be specified using the following boolean SQL query.
\begin{verbatim}
select not exists (select 1
      	      	     from	  Skill	s
      	      	     where  not exists (select 1
      	      	      	      	          from   personSkill ps
      	      	      	      	          where  ps.skill = s.skill));
\end{verbatim}

\end{example}


\begin{enumerate}[resume]

\item  Each Manager manages at least two people. (5 Points)

\item  Some person has a salary that is strictly lower than the salary of each of his or her managers. (5 Points)

\item  Each employee and his or her managers work for the same company. (5 Points)


\end{enumerate}


\end{document}
